---
title: "Actuarial Present Value and IRR Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actuarial Present Value and IRR Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(abslife)
```

This vignette provides a clear example of how to calculate the Actuarial Present
Value (APV) and risk-adjusted Internal Rate of Return (IRR) for a loan portfolio
subject to competing risks (Default and Pre-payment).

We use the built-in `aloans` dataset from the `abslife` package, which contains
left-truncated and right-censored loan performance records.

## Step 1: Hazard Estimation with Competing Risks

First, we estimate the hazard rates for the competing risks (Default
vs. Pre-payment) conditioning on left-truncation. We focus on deep subprime
loans and specify `support_lifetime_rv` up to the loan term of 72 months to
cover the entire potential duration.

```{r hazard-estimation}
data(aloans)

orig_term <- 72
cur_age <- 7

# Estimate hazards for the competing risk events
hazards <-
  with(subset(aloans, risk.cat == "deep_subprime"),
       estimate_hazard(lifetime = Z,
                       trunc_time = Y,
                       event_type = ifelse(D == 1, "Default", "Pre-payment"),
                       censoring = C,
                       ci_level = 0.95,
                       carry_hazard = TRUE))
```

## Step 2: Compute CDF and Densities

Next, we convert these hazard rates into cumulative distribution functions (CDF)
and sub-density values using `calc_cdf`.

```{r calc-cdf}
prbs <- calc_cdf(hazards)
```

## Step 3: Setup Recovery Curve

We define a recovery curve mapping the recovery percentage to each remaining
month of the loan's term. In this example, we construct a dummy recovery curve
showing declining recovery over time.

```{r recovery-curve}
rem_months <- cur_age:(orig_term - 1)
recov_curve <- data.frame(
  Month = rem_months,
  Recovery = seq(0.40, 0.15, length.out = length(rem_months))
)
```

## Step 4: Calculate Actuarial Present Value (APV)

We set the loan parameters (original amount, monthly payment, and reference
market rate) and compute the APV:

```{r calculate-apv}
orig_loan_amt <- 50000
mon_pmt <- 1031.84
ref_rate <- 0.05 # Annual market discount rate

apv_res <- calculate_apv(
  cur_age = cur_age,
  orig_term = orig_term,
  orig_loan_amt = orig_loan_amt,
  mon_pmt = mon_pmt,
  ref_rate = ref_rate,
  recov_curve = recov_curve,
  prbs = prbs,
  orig_apy = 0.15
)
```

* Estimated APV: `r sprintf("$%.2f (%.2f)", apv_res$APV, apv_res$SD)`



## Step 5: Solve for Risk-Adjusted IRR

Assuming a current loan price/balance of `$37,719.47`, we calculate the
risk-adjusted monthly and annualized Internal Rate of Return (IRR) that equates
the loan price to its expected present value:

```{r solve-irr}
cur_bal <- 37719.47

irr_res <- solve_irr(
  cur_bal = cur_bal,
  cur_age = cur_age,
  orig_term = orig_term,
  orig_loan_amt = orig_loan_amt,
  mon_pmt = mon_pmt,
  recov_curve = recov_curve,
  prbs = prbs,
  orig_apy = 0.15
)
```

* Calculated monthly rate: `r sprintf("%.6f (%.4f%%)", irr_res$monthly_rate,
  irr_res$monthly_rate * 100)`
* Calculated annualized rate: `r sprintf("%.6f (%.4f%%)", irr_res$annualized_rate, irr_res$annualized_rate * 100)`
